Detection and partial characterisation of stellar multiplicity

### 2018-09-03 article/ms.tex
changes

\title{Detection and partial characterisation of stellar multiplicity with Gaia}

@@

\author{Rosemary Mardling}
\author{John Lattanzio}

\author[0000-0003-2866-9403]{David W. Hogg}
\affiliation{Center for Cosmology and Particle Physics, Department of Physics,
    		 New York University, 
		  	 726 Broadway, New York, NY 10003, USA}
\affiliation{Center for Data Science, 
			 New York University, 60 Fifth Ave, New York, NY 10011, USA}
\affiliation{Max-Planck-Institut f\"ur Astronomie, 
			 K\"onigstuhl 17, D-69117 Heidelberg}
\affiliation{Flatiron Institute,
			 162 Fifth Ave, New York, NY 10010, USA}

\author[0000-0003-0872-7098]{Adrian M. Price-Whelan}
\affiliation{Department of Astrophysical Sciences, 
			 Princeton University, 
			 Princeton, NJ 08544, USA}

@@

Some sources in \Gaia\ DR2 do not have reported radial velocities even though they are
apparently bright enough and within a suitable \bp\ - \rp\ colour range. The two most
likely explanations are that either the \Gaia\ team identified the source to be a
double-lined spectroscopic binary, or the radial velocity error was larger than 20\,km\,s$^{-1}$,
and so the source radial velocity was removed. While it may be appealing to use the \emph{lack}
of radial velocity measurement as an indicator of stellar multiplicity, we show in Appendix A
that such an approach is problematic due to completeness, and other complications arising from
the initial \Gaia\ source catalog. However, in the following sections we show that for 
sources that do not have reported radial velocities -- and it has an apparent magnitude 
and colour that suggests it ought to have a radial velocity reported -- the astrometric 
jitter is often sufficient to identify stellar multiples.

% todo:  Figure showing distribution of astrometric unit weight error.

@@

We restrict our work to the brightest $\sim{}10^7$ sources in \Gaia\ DR2.
About 70\% of sources in this subset have reported radial velocities, and
this subset is relatively close to the Sun. Specifically, we executed the
following Astronomical Data Query Language \citep[\texttt{ADQL};][]{someone} 
query on \ESA\ \Gaia\ archive \citep{esa_gaia_archive}:

\begin{minted}[style=friendly]{postgresql}
SELECT *
  FROM gaiadr2.gaia_source
 WHERE radial_velocity IS NOT NULL
    OR phot_rp_mean_mag <= 13
\end{minted}
\noindent{}This query returned \NumberOfSourcesInSubset\ sources.

@@

We assume that for a sample of sources with similar properties:
\begin{itemize}
	\item \bp\ - \rp\ colours;
	\item apparent \rp\ magnitude; and
	\item absolute \rp\ magnitude,
\end{itemize}
\noindent{}the distribution of jitter (in astrometry or radial velocity)
is a mixture of two components:

@@

We further assume that the jitter for a single star will change depending on the
source properties. For example, the mean radial velocity jitter of single stars 
with $\teff \approx 6000\,\textrm{K}$ will be higher than the jitter of single stars with
$\teff \approx 5000\,\textrm{K}$, all else being equal. For these reasons, in order to evaluate whether 
a source is more likely to be a single star or a stellar multiple, we must consider 
the jitter in context with other sources that have a similar properties (\bp\ - \rp\ colour,
apparent magnitude, and absolute magnitude).\footnote{The magnitude in any band would be sufficient. Throughout this work we chose the \rp\ band for radial velocity data because the wavelength of the \rp\ band most closely overlaps with that of the Radial Velocity Spectrograph on board \Gaia. For similar reasons we use the $G$ band magnitudes for astrometry data.}
However, we cannot reliably describe a quantitative
relationship for \emph{how} exactly the jitter of a single star is
expected to vary given some source properties. For these reasons
we chose to adopt a flexible non-parametric model for stellar multiplicity, primarily
conditioned on the assumption that if we select enough sources with \emph{similar}
properties then within that subset of similar stars there exists \emph{some} single stars 
and \emph{some} stellar multiples -- even if we do not necessarily know \emph{a priori}
which sources are single stars, and which are stellar multiples.

@@

@@

\begin{figure}
%	\includegraphics[width=0.4\textwidth]{../figures/sb2_histograms.pdf}
	\includegraphics[width=0.4\textwidth]{../figures/todo.png}
    \caption{Distribution of 
    			(a) astrometric unit weight error (see Eq. \ref{eq:astrometric_unit_weight_error}),
			%$u = \left(\chi_{al}^2/(N_{al} - 5)\right)^{1/2}$
			%		where $\chi_{al}^2$ and $N_{al}$ is the astrometric goodness-of-fit and number 
			%		of observations in the along-scan direction, respectively (columns 
			%			\texttt{astrometric\_chi2\_al} and \texttt{astrometric\_n\_obs\_al}), 
				(b) photometric \bp\ - \rp\ excess factor, 				
	 				and 
				(c) photometric \rp\ variability (see Eq. \ref{eq:photometric_variability})
			of candidate SB2 type systems relative to a control sample of the same size.}
    \label{fig:sb2_histograms}
\end{figure}

@@

%\begin{eqnarray}
%	\textrm{Amp} = \log_{10}\left(\sqrt{N_{obs,al}}\frac{\sigma_{\langle{}f_G\rangle}}{\langle{}f_G\rangle}\right) \label{eq:photometric_variability}
%\end{eqnarray}

@@

\section{Radial velocity completeness as a function of \Gaia\ source properties}

@@

>>>>

\title{Detection and partial characterisation of stellar multiplicity}

@@

\author{Rosemary Mardling}
\affiliation{School of Physics \& Astronomy, 
			 Monash University,
			 Wellington Rd, Clayton 3800, Victoria, Australia}

\author{John Lattanzio}
\affiliation{School of Physics \& Astronomy, 
			 Monash University,
			 Wellington Rd, Clayton 3800, Victoria, Australia}

%\author[0000-0003-2866-9403]{David W. Hogg}
%\affiliation{Center for Cosmology and Particle Physics, Department of Physics,
%    		 New York University, 
%		  	 726 Broadway, New York, NY 10003, USA}
%\affiliation{Center for Data Science, 
%			 New York University, 60 Fifth Ave, New York, NY 10011, USA}
%\affiliation{Max-Planck-Institut f\"ur Astronomie, 
%			 K\"onigstuhl 17, D-69117 Heidelberg}
%\affiliation{Flatiron Institute,
%			 162 Fifth Ave, New York, NY 10010, USA}

%\author[0000-0003-0872-7098]{Adrian M. Price-Whelan}
%\affiliation{Department of Astrophysical Sciences, 
%			 Princeton University, 
%			 Princeton, NJ 08544, USA}

@@

Some sources in \Gaia\ DR2 do not have reported radial velocities even though they are
apparently bright enough and within a suitable \bp\ - \rp\ colour range. The two most
likely explanations are that either the \Gaia\ team identified the source to be a
double-lined spectroscopic binary, or the radial velocity error $\sigma_{\textrm{V}_\textrm{R}}^\textrm{MTA}$ exceeded 20\,km\,s$^{-1}$ (irrespective of the number of transits),
and so the source radial velocity was removed. While it may be appealing to use the \emph{lack}
of radial velocity measurement as an indicator of stellar multiplicity, we show in Appendix \ref{sec:appendix-missing-rvs}
that such an approach is problematic due to completeness, and other complications arising from
the initial \Gaia\ source catalog. However, in later sections we also show that for 
sources without reported radial velocities -- and those sources have apparent magnitudes
and colours suggesting it ought to have a radial velocity measurement -- the astrometric 
jitter is often sufficient to identify stellar multiples (see also Figure \ref{fig:sb2_histograms}).

@@

We restrict our work to the brightest $\sim{}10^7$ sources in \Gaia\ DR2.
Specifically, we executed the
following Astronomical Data Query Language \citep[\texttt{ADQL};][]{someone} 
query on \ESA\ \Gaia\ archive \citep{esa_gaia_archive}:

\begin{minted}[style=friendly]{postgresql}
SELECT *
  FROM gaiadr2.gaia_source
 WHERE radial_velocity IS NOT NULL
    OR phot_rp_mean_mag <= 13
\end{minted}
\noindent{}This query returned \NumberOfSourcesInSubset\ sources, of which
about 70\% have radial velocities in \Gaia\ DR2.

@@

We assume that for a sample of sources with a similar \bp\ - \rp\ colour,
apparent magnitude\footnote{The magnitude in any band is sufficient. Throughout this work we chose the \rp\ band for radial velocity data because the wavelength of the \rp\ band most closely overlaps with that of the Radial Velocity Spectrograph on board \Gaia. For similar reasons we use the G-band magnitudes for astrometry data.}, and absolute magnitude, 
the distribution of jitter (in astrometry or radial velocity)
is a mixture of two components:

@@

We further assume that the jitter for a single star will change depending on the
source properties. For example, the mean radial velocity jitter of single stars 
with $\teff \approx 6000\,\textrm{K}$ will be higher than the jitter of single stars with
$\teff \approx 5000\,\textrm{K}$, all else being equal. For these reasons, in order to evaluate whether 
a source is more likely to be a single star or a stellar multiple, we must consider 
the jitter in context with other sources that have a similar properties (\bp\ - \rp\ colour,
apparent magnitude, and absolute magnitude).

However, we cannot reliably describe a quantitative
relationship for \emph{how} exactly the jitter of a single star is
expected to vary given some source properties. For these reasons
we chose to adopt a flexible non-parametric model for stellar multiplicity, primarily
conditioned on the assumption that if we select enough sources with \emph{similar}
properties then within that subset of similar stars there exists \emph{some} single stars 
and \emph{some} stellar multiples -- even if we do not necessarily know \emph{a priori}
which sources are single stars, and which are stellar multiples.

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{../figures/todo.png}
%	\includegraphics[width=\textwidth]{figures/tau_single_hrd.png}
    \caption{Two-dimensional histogram of \todo{X} showing the mean single star probability $\tau$ per bin. The three panels show probabilities calculated from different sources of information: (a) using only radial velocity jitter; (b) using only astrometric jitter; and (c) using the joint information of radial velocity and astrometric jitter. All panels share the same colour range.}
    \label{fig:tau_single_hrd_v5}
\end{figure*}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.35\textwidth]{../figures/sb2_histograms.pdf}
%	\includegraphics[width=0.4\textwidth]{../figures/todo.png}
    \caption{Distribution of 
    			(a) astrometric unit weight error (see Eq. \ref{eq:astrometric_unit_weight_error}),
			%$u = \left(\chi_{al}^2/(N_{al} - 5)\right)^{1/2}$
			%		where $\chi_{al}^2$ and $N_{al}$ is the astrometric goodness-of-fit and number 
			%		of observations in the along-scan direction, respectively (columns 
			%			\texttt{astrometric\_chi2\_al} and \texttt{astrometric\_n\_obs\_al}), 
				(b) photometric \bp\ - \rp\ excess factor, 				
	 				and 
				(c) photometric \rp\ variability (see Eq. \ref{eq:photometric_variability})
			of sources that lack radial velocities but their properties suggest they should have
			a radial velocity measurement (red; candidate SB2 systems), and an equal-size control sample of sources
			with similar \bp\ - \rp\ colours, apparent magnitudes, absolute magnitudes,
			which \emph{do} have radial velocity measurements (grey). It is clear that the
			candidate SB2 systems have systematically higher astrometric unit weight error
			than the control sample, and a larger variance in photometric \bp\ - \rp\ excess
			factor and photometric variability, both consistent with a secondary source.}
    \label{fig:sb2_histograms}
\end{figure}

@@

@@

\section{Radial velocity completeness as a function of \Gaia\ source properties} \label{sec:appendix-missing-rvs}

@@

\todo{\begin{eqnarray}
	\textrm{Amp} = \log_{10}\left(\sqrt{N_{obs,al}}\frac{\sigma_{\langle{}f_G\rangle}}{\langle{}f_G\rangle}\right) \label{eq:photometric_variability}
\end{eqnarray}}
